Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2020-01-29 rnni_convexity.tex
add intro partition lattice + refer to it earlier

The rest of this paper is organised as follows.
In Section~\ref{section:algorithms} we introduce three algorithms for exploring the $\rnni$ graph.
We use these algorithms in Section~\ref{section:geometry} to establish several geometrical properties of this graph, such as its diameter and radius.
Furthermore, we will establish in Section~\ref{section:caterpillar_convex} that a certain subspace of $\rnni$ is convex and design a polynomial-time algorithm for computing $\rnni$-distances in that subspace.
This result suggests that the complexity of computing distances in $\rnni$ is different from that in the classical $\nni$ space.
We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest weaker versions as plausible alternatives.
We will finish the paper with a discussion and open problems.

@@

The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
Those algorithms will, for example, enable us to prove that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} defined as $\min\limits_T \max\limits_R d(T,R)$.
Next, we will consider the so-called Split Property \autocite{Gavryushkin2018-ol}, that is, every shared split of taxa is maintained along every shortest path.
Recall that a \emph{split} is a bipartition of the set of taxa obtained by deleting an edge of the tree $T$.
Splits are denoted by $A|B$.
We will give a counterexample to show that $\rnni$ does not have the Split Property and consider a variant, the Weak Split Property, which says that every shared split of taxa is maintained along a shortest path.
We will also consider the so-called Cluster Property as another weak version of the Split Property (see Conjecture~\ref{conjecture:cluster_theorem}).

@@

\subsection{Partition lattice}

>>>>

The rest of this paper is organised as follows.
In Section~\ref{section:algorithms} we introduce three algorithms for exploring the $\rnni$ graph.
We use these algorithms in Section~\ref{section:geometry} to establish several geometrical properties of this graph, such as its diameter and radius.
Furthermore, we will establish in Section~\ref{section:caterpillar_convex} that a certain subspace of $\rnni$ is convex and design a polynomial-time algorithm for computing $\rnni$-distances in that subspace.
This result suggests that the complexity of computing distances in $\rnni$ is different from that in the classical $\nni$ space.
We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest weaker versions as plausible alternatives.
In Section~\ref{section:partition_lattice} we establish a connection between the $\rnni$ graph and a classical algebraic structure, the partition lattice.
We will finish the paper with a discussion and open problems.

@@

The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
Those algorithms will, for example, enable us to prove that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} defined as $\min\limits_T \max\limits_R d(T,R)$.
Next, we will consider the so-called Split Property \autocite{Gavryushkin2018-ol}, that is, every shared split of taxa is maintained along every shortest path.
Recall that a \emph{split} is a bipartition of the set of taxa obtained by deleting an edge of the tree $T$.
Splits are denoted by $A|B$.
We will give a counterexample to show that $\rnni$ does not have the Split Property and consider a variant, the Weak Split Property, which says that every shared split of taxa is maintained along a shortest path.
We will also consider the so-called Cluster Property as another weak version of the Split Property (see Conjecture~\ref{conjecture:cluster_theorem}).
In the final part of this geometry section we will discuss the relation between the $\rnni$ graph and the so-called partition lattice, a well known algebraic structure.

@@

\subsection{Partition lattice}
\label{section:partition_lattice}

In this section we establish a connection between the $\rnni$ graph and a well known algebraic structure, the partition lattice.
The relation between these two provides a new direction for further research and leads to a different view on problems of either structure.
